---
title: "P8105_HW2_jao2195"
output: github_document
---

### Problem 1.1 - Install Packages / Load Libraries

This problem focuses the use of inline R code, plotting, and the behavior of ggplot for variables of different types.
Use the code below to download the a package containing the early_january_weather dataset:

install.packages("moderndive").

You only need to run this command once to install the package, and you can do so directly in the console. This code shouldn’t be executed by your RMarkdown file.

Load the moderndive library, and use the following code to load the early_january_weather dataset:

```{r, message=FALSE}
#1. Installing Packages and Libraries 

# Packages already installed, previously. 
# Loading Libraries
library(moderndive)
library(ggplot2)
library(tidyverse)

#2. Loading early_january_weather dataset from moderndive package.
data(early_january_weather)

#3. Looking at help function to learn more about the variables in the dataset. 
help(early_january_weather)

```

### Problem 1.2 - Inline R code Dataset Discussion Short Description

Write a short description of the dataset using inline R code accessing the dataset help file can be informative. In your discussion, please include:

*the variables in this dataset, including names / values of important variables

*the size of the dataset (using nrow and ncol)

*the mean temperature

################################################################################
### My Short Description of Early January Weather Dataset (1.2)

The Early January Weather data (early_january_weather) is a dataset that contains hourly meterological data for Laguardia (LGA), John F. Kennedy (JFK) , and Newark (EWR) Airports in the New York City area. This data is a subset of the nycflights13 dataset in R.  This dataset contains `r ncol(early_january_weather)` variables: `r colnames(early_january_weather)` and `r nrow(early_january_weather)` observations. Values of important variables in the dataset including but not limited to: Average Preciptation, which was `r round(mean(pull(early_january_weather, precip)))` inches. Maximum visibility in early January was `r max(pull(early_january_weather,visib))` miles. The mean (average) temperature in early January 2013 was `r round(mean(pull(early_january_weather, temp)))` degrees Fahrenheit !

################################################################################
### Problem 1.3 - Creating Scatterplot of Humidity/Temperature Over Time 

Make a scatterplot of temp (y) vs time_hour (x); color points using the humid variable (adding color = ... inside of aes in your ggplot code should help). Describe patterns that are apparent in this plot.
Export your scatterplot to your project directory using ggsave.


```{r, message=FALSE}
#1. Creating Scatterplot, storing in myplot object.
myplot <-
  ggplot(early_january_weather) + 
  geom_point(aes(x= time_hour, y = temp, col= humid)) + 
  labs(x="Time", y="Temperature (Fahrenheit, F)", col="Relative Humidity", 
       title="Humidity/Temperature Over Time in Early January 2013 ") 

#2. Printing myplot object to rendered document.
myplot

#3. Saving .png file version to plot to working directory (wd).
ggsave("Temp_Humid_Scatterplot_Problem 1.png") 

```

################################################################################
### My Short Description of the Scatterplot (1.3)

During early January 2013, temperature ranged from `r round(min(pull(early_january_weather, temp)))` and `r round(max(pull(early_january_weather, temp)))`degrees Fahrenheit. 
Temperature fluctuated multiple times but had general upward tread (increasing temperature with time). 
In addition, humidity also increased with time, with humidity being the highest between January 11 - January 15th. 
In general, as temperature increased, relative humidity increased.

################################################################################
################################################################################

### Problem 2.1 

This problem is intended to emphasize variable types and introduce coercion; some awareness of how R treats numeric, character, and factor variables is necessary for working with these data types in practice.

Create a data frame comprised of:

a random sample of size 10 from a standard Normal distribution
a logical vector indicating whether elements of the sample are greater than 0
a character vector of length 10
a factor vector of length 10, with 3 different factor “levels”
Try to take the mean of each variable in your dataframe. What works and what doesn’t?

Hint: for now, to take the mean of a variable in a dataframe, you need to pull the variable out of the dataframe. Try loading the tidyverse and using the pull function.


```{r}


jennsdataframe <- tibble (
                    rand_sample = rnorm(10,0,1), 
                    logical_vec = rand_sample>0,
                    char_vec = c("Columbia","University","New","York",
                                 "Manhattan","Brooklyn","Queens","Staten Island",
                                 "Bronx","Long Island"), 
                    factor_vec = factor(c("High", "Medium","Low", "Low", 
                                   "High","Medium", "Low", "Low", 
                                   "Medium", "High"), levels = c("Low", "Medium", "High")))
  
#Mean of random sample, rand_samp:                  
rand <- mean(pull(jennsdataframe, rand_sample))

#Mean of logical vector, logical_vec:  
logic <- mean(pull(jennsdataframe, logical_vec))

#Mean of character vector, char_vec:  
char <- mean(pull(jennsdataframe, char_vec))

#Mean of factor vector, factor_vec:  
factor <- mean(pull(jennsdataframe, factor_vec))


```
We can see that for:

1) Random Sample - A mean of  **`r rand`** is generated for our random variable vector.A mean number is returned. This works.
2) Logical Vector - A mean of  **`r logic`** is generated for our logical variable vector. A mean number is returned. This works. R can translate logical to numeric, as seen in error messages of 3 & 4 below.
3) Character Vector - A mean of  **`r char`** is generated for our character variable vector. NA returned because Warning: argument is not numeric or logical.This does not work. Mean() only works on a numeric or logical type.
4) Factor Vector  - A mean of  **`r factor`** is generated for our factor variable vector. NA returned because Warning: argument is not numeric or logical.This does not work. Mean() only works on a numeric or logical type.


#################################
### Problem 2.2 
In some cases, you can explicitly convert variables from one type to another. Write a code chunk that applies the as.numeric function to the logical, character, and factor variables (please show this chunk but not the output). What happens, and why? Does this help explain what happens when you try to take the mean?

```{r, warning=FALSE}

#Mean of logical vector, logical_vec:  
logic_num <- mean(as.numeric(pull(jennsdataframe, logical_vec)))

#Mean of character vector, char_vec:  
char_num <- mean(as.numeric(pull(jennsdataframe, char_vec)))

#Mean of factor vector, factor_vec:  
factor_num <- mean(as.numeric(pull(jennsdataframe, factor_vec)))
```

We can see that for the new as.numeric() coerced variables:

1) Logical Vector - A mean of  **`r logic_num`** is generated for our vector. A mean number is returned. This happens because the logical values of TRUE and FALSE are converted to numerical 1s and 0s.  Yes this does explain what happens when we are trying to take the mean as the mean value is the average of the 0s and 1s in the vector. 
2) Character Vector - A mean of  **`r char_num`** is generated for our vector. NA returned because Warning: NAs introduced by coercion. This happened because you cannot convert a character into a number. This does explain what happened when were trying to take the mean as we cannot compute a mean on characters. 
3) Factor Vector  - A mean of  **`r factor_num`** is generated for our vector. A mean number is returned. This happens because the levels of the factors (high, med, low) are converted to numerical values (1,2,3) for each of the levels. Yes, this does explain what happens when we are trying to take the mean, as the mean value is the average of the levels of the factors in the vector. 

###############################
####   End Of Homework 1   ####
###############################